---
title: "US Health & Nutrition Trends (1999–2023)"
author: "Sam Martano"
format:
  html:
    theme: cosmo
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| include: false
library(tidyverse)
library(plotly)
```

::: {.panel-tabset}

## US Adult Obesity Prevalence (1999–2023)

This visualization shows the trend in age-adjusted obesity prevalence ($\text{BMI} \ge 30$) among adults aged 20 and older in the United States from 1999 through 2023, based on data collected by the National Health and Nutrition Examination Survey (NHANES).

```{r}
# CDC NHANES age-adjusted adult obesity prevalence across survey cycles
obesity_trends <- tibble(
  cycle = c(
    "1999–2000", "2001–2002", "2003–2004", "2005–2006",
    "2007–2008", "2009–2010", "2011–2012", "2013–2014",
    "2015–2016", "2017–2018", "2017–2020", "2021–2023"
  ),
  year = c(1999.5, 2001.5, 2003.5, 2005.5, 2007.5, 2009.5, 2011.5, 2013.5, 2015.5, 2017.5, 2018.5, 2022.5),
  prevalence = c(30.5, 30.5, 32.2, 34.3, 33.7, 35.7, 34.9, 37.7, 39.6, 42.4, 41.9, 40.3)
)

# Build plot with ggplot2
p <- ggplot(obesity_trends, aes(x = year, y = prevalence)) +
  geom_line(aes(group = 1), color = "#2b6cb0", linewidth = 1) +
  geom_point(
    aes(text = paste0("Cycle: ", cycle, "<br>Prevalence: ", prevalence, "%")),
    color = "#2b6cb0",
    size = 2.5
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, "%"),
    limits = c(25, 45)
  ) +
  scale_x_continuous(
    breaks = seq(2000, 2022, by = 4),
    limits = c(1998, 2024)
  ) +
  labs(
    title = "Adult Obesity Rate Over Time (NHANES 1999–2023)",
    x = "Survey Period",
    y = "Obesity Prevalence (%)"
  ) +
  theme_minimal()

# Convert to interactive plotly widget
ggplotly(p, tooltip = "text")
```

Between 1999–2000 and 2017–2018, adult obesity prevalence rose steadily from 30.5% to over 42%, before leveling off near 40% in the most recent 2021–2023 cycle.

## Diet-Related Chronic Diseases (1999–2023)

This visualization tracks the age-adjusted prevalence of major chronic health conditions in the U.S. that are strongly influenced by diet, nutrition, and food quality, based on examination data from CDC NHANES.

```{r}
# CDC NHANES age-adjusted chronic disease prevalence across survey cycles
cycles <- c(
  "1999–2000", "2001–2002", "2003–2004", "2005–2006", "2007–2008",
  "2009–2010", "2011–2012", "2013–2014", "2015–2016", "2017–2018",
  "2017–2020", "2021–2023"
)
years <- c(1999.5, 2001.5, 2003.5, 2005.5, 2007.5, 2009.5, 2011.5, 2013.5, 2015.5, 2017.5, 2018.5, 2022.5)

chronic_trends <- tibble(
  cycle = rep(cycles, 4),
  year = rep(years, 4),
  disease = c(
    rep("Hypertension", 12),
    rep("Metabolic Syndrome", 12),
    rep("High Cholesterol", 12),
    rep("Diabetes", 12)
  ),
  rate = c(
    # Hypertension (BP >= 130/80 or medication)
    47.0, 46.0, 44.9, 43.8, 43.2, 42.1, 43.9, 41.7, 43.6, 45.4, 45.1, 45.5,
    # Metabolic Syndrome (cluster of cardiometabolic risk factors)
    28.2, 29.0, 29.8, 30.7, 31.5, 31.9, 33.2, 34.6, 35.8, 37.1, 37.5, 36.8,
    # High Cholesterol (total cholesterol >= 240 mg/dL)
    18.3, 17.2, 16.5, 15.7, 14.6, 13.4, 12.9, 12.1, 12.4, 11.5, 10.8, 10.5,
    # Diabetes (total diagnosed and undiagnosed)
    9.7, 10.4, 10.9, 11.5, 11.9, 12.2, 12.5, 13.0, 14.9, 14.8, 14.8, 14.3
  )
)

# Build plot with ggplot2
p_chronic <- ggplot(chronic_trends, aes(x = year, y = rate, color = disease)) +
  geom_line(aes(group = disease), linewidth = 1) +
  geom_point(
    aes(text = paste0("Condition: ", disease, "<br>Cycle: ", cycle, "<br>Prevalence: ", rate, "%")),
    size = 2.5
  ) +
  scale_color_manual(
    values = c(
      "Hypertension" = "#e65100",        # deep orange
      "Metabolic Syndrome" = "#0288d1",  # blue
      "High Cholesterol" = "#c2185b",    # berry/magenta
      "Diabetes" = "#7b1fa2"             # purple
    )
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, "%"),
    limits = c(5, 55)
  ) +
  scale_x_continuous(
    breaks = seq(2000, 2022, by = 4),
    limits = c(1998, 2024)
  ) +
  labs(
    title = "Prevalence of Diet-Related Chronic Diseases (NHANES 1999–2023)",
    x = "Survey Period",
    y = "Age-Adjusted Prevalence (%)",
    color = "Condition"
  ) +
  theme_minimal()

# Convert to interactive plotly widget
ggplotly(p_chronic, tooltip = "text")
```

### About These Diseases and Their Links to Food

::: {.callout-note collapse="true" icon=false}
## Hypertension (High Blood Pressure)
Defined as blood pressure $\ge 130/80\text{ mmHg}$ or taking blood pressure medication. Excessive sodium (salt) from processed and restaurant foods, combined with diets low in potassium and magnesium from fresh fruits and vegetables, is a key driver.
:::

::: {.callout-note collapse="true" icon=false}
## Metabolic Syndrome
A cluster of at least three concurrent cardiometabolic risk factors: abdominal obesity (large waist circumference), high triglycerides, low HDL ("good") cholesterol, high blood pressure, and elevated fasting blood glucose. Heavily promoted by diets high in ultra-processed foods and refined carbohydrates.
:::

::: {.callout-note collapse="true" icon=false}
## High Total Cholesterol (Hypercholesterolemia)
Defined as blood total cholesterol $\ge 240\text{ mg/dL}$. Elevated by diets high in saturated fats and industrial trans fats. Its long-term decline over this period is largely credited to widespread clinical screening and statin medication use.
:::

::: {.callout-note collapse="true" icon=false}
## Diabetes (Total Type 2 & Type 1)
Defined by elevated blood glucose ($\text{HbA1c} \ge 6.5\%$ or fasting glucose $\ge 126\text{ mg/dL}$), with Type 2 accounting for over 90% of cases. Strongly linked to sugar-sweetened beverages, ultra-processed food intake, and excess caloric consumption that lead to insulin resistance.
:::

:::

## Food Consumption Trends (Ultra-Processed vs. Unprocessed)

This visualization tracks the proportion of daily caloric intake derived from **ultra-processed foods** (red) compared to **unprocessed or minimally processed foods** (green) across all survey cycles from 1999 through 2023 based on NHANES dietary recall data.

```{r}
# NHANES dietary energy intake (% of total kcal) categorized by NOVA food processing level
food_trends <- tibble(
  cycle = rep(
    c(
      "1999–2000", "2001–2002", "2003–2004", "2005–2006", "2007–2008",
      "2009–2010", "2011–2012", "2013–2014", "2015–2016", "2017–2018",
      "2017–2020", "2021–2023"
    ),
    2
  ),
  year = rep(
    c(1999.5, 2001.5, 2003.5, 2005.5, 2007.5, 2009.5, 2011.5, 2013.5, 2015.5, 2017.5, 2018.5, 2022.5),
    2
  ),
  food_type = c(
    rep("Ultra-Processed Food", 12),
    rep("Unprocessed Food", 12)
  ),
  rate = c(
    52.8, 53.5, 54.0, 54.8, 55.2, 55.6, 56.1, 56.4, 56.7, 57.0, 56.5, 53.0,
    33.5, 32.7, 32.1, 31.2, 30.5, 29.8, 29.1, 28.5, 27.9, 27.4, 28.0, 31.2
  )
)

# Build plot with red for processed and green for unprocessed
p_food <- ggplot(food_trends, aes(x = year, y = rate, color = food_type)) +
  geom_line(aes(group = food_type), linewidth = 1) +
  geom_point(
    aes(text = paste0("Type: ", food_type, "<br>Cycle: ", cycle, "<br>Caloric Share: ", rate, "%")),
    size = 2.5
  ) +
  scale_color_manual(
    values = c(
      "Ultra-Processed Food" = "#d9534f", # red
      "Unprocessed Food" = "#2e7d32"       # green
    )
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, "%"),
    limits = c(20, 65)
  ) +
  scale_x_continuous(
    breaks = seq(2000, 2022, by = 4),
    limits = c(1998, 2024)
  ) +
  labs(
    title = "Ultra-Processed vs. Unprocessed Food Consumption (NHANES 1999–2023)",
    x = "Survey Period",
    y = "Share of Daily Energy Intake (%)",
    color = "Food Category"
  ) +
  theme_minimal()

# Convert to interactive plotly widget
ggplotly(p_food, tooltip = "text")
```

From 1999 to 2018, daily caloric intake from ultra-processed food (red) rose from 52.8% to 57.0%, while unprocessed food (green) dropped from 33.5% to 27.4%, followed by a post-pandemic shift toward whole foods in 2021–2023.
