---
title: "US Health & Nutrition Trends (1999–2023)"
author: "Sam Martano"
format:
  html:
    theme: cosmo
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| include: false
library(tidyverse)
library(plotly)
library(NHANES)
```

<style>
body {
  background-color: #f2f7f4;
  background-image: url("data:image/svg+xml,%3Csvg xmlns='http://www.w3.org/2000/svg' width='160' height='160' viewBox='0 0 160 160' fill='none' stroke='%232e7d32' stroke-width='1.4' stroke-linecap='round' stroke-linejoin='round' opacity='0.08'%3E%3Cpath d='M35 32 C26 32 20 39 20 49 C20 62 35 72 35 72 C35 72 50 62 50 49 C50 39 44 32 35 32 Z'/%3E%3Cpath d='M35 32 Q34 24 38 20'/%3E%3Cpath d='M36 26 Q43 22 46 26 Q42 29 36 26'/%3E%3Cpath d='M115 28 C122 25 128 30 125 35 L114 62 C113 64 110 64 109 62 L103 40 C101 34 107 27 115 28 Z'/%3E%3Cpath d='M116 27 L116 16 M118 27 L125 18 M114 27 L107 19'/%3E%3Cpath d='M32 110 C22 96 36 88 46 92 C46 102 41 116 32 110 Z'/%3E%3Cpath d='M32 110 C30 116 26 122 22 125'/%3E%3Cpath d='M46 92 C54 84 66 94 60 102 C54 102 48 98 46 92 Z'/%3E%3Cpath d='M120 105 C120 97 112 92 105 97 C98 92 90 97 90 105 C90 117 105 127 105 127 C105 127 120 117 120 105 Z'/%3E%3Cpath d='M96 107 L101 107 L104 101 L107 113 L110 107 L114 107'/%3E%3Cpath d='M78 68 L82 88'/%3E%3Cpath d='M78 72 C74 70 73 66 77 64 C80 66 80 70 78 72 Z'/%3E%3Cpath d='M80 74 C84 72 85 68 81 66 C78 68 78 72 80 74 Z'/%3E%3Cpath d='M79 80 C75 78 74 74 78 72 C81 74 81 78 79 80 Z'/%3E%3Cpath d='M81 82 C85 80 86 76 82 74 C79 76 79 80 81 82 Z'/%3E%3C/svg%3E");
  background-repeat: repeat;
  background-size: 160px 160px;
}

.hero-card {
  background: linear-gradient(135deg, #e8f5e9 0%, #e0ece4 50%, #edf2f7 100%);
  border-radius: 12px;
}

.hero-card h3 {
  color: #1b4332;
  font-weight: 700;
  margin-top: 0;
  margin-bottom: 0.75rem;
}

.tab-content {
  background-color: rgba(255, 255, 255, 0.9);
  border-radius: 0 0 12px 12px;
  padding: 1.5rem;
  box-shadow: 0 2px 6px rgba(0, 0, 0, 0.04);
}
</style>

::: {.hero-card .card .p-4 .mb-4 .border-0 .shadow-sm}
### Exploring 20 Years of American Nutrition & Health

Diet and nutrition play a central role in determining long-term health outcomes. When diets are heavily reliant on ultra-processed foods—which are often high in added sugars, unhealthy fats, and sodium—they contribute to excess weight gain and substantially elevate the risk of chronic conditions such as type 2 diabetes and hypertension. Conversely, diets rich in whole, minimally processed foods provide essential nutrients and fiber that support a healthy weight and protect metabolic function. The visualizations below draw on data from the National Health and Nutrition Examination Survey (NHANES) to examine how changes in American dietary habits connect to trends in body mass index (BMI), adult obesity, and chronic disease over the last two decades.
:::

::: {.panel-tabset}

## Adult BMI by Height and Weight

This visualization maps adult survey participants from the NHANES dataset across height and weight. Points in **green** indicate a healthy weight ($\text{BMI} < 25$), while points in **red** indicate overweight or obesity ($\text{BMI} \ge 25$).

```{r}
#| fig-width: 10
#| fig-height: 7
# Filter adult participants with recorded measurements, removing extreme outliers
nhanes_bmi <- NHANES |>
  distinct(ID, .keep_all = TRUE) |>
  filter(Age >= 20, !is.na(Height), !is.na(Weight), !is.na(BMI)) |>
  mutate(
    height_in = Height / 2.54,
    weight_lbs = Weight * 2.20462,
    status = if_else(BMI >= 25, "Overweight or Obese", "Healthy"),
    status = factor(status, levels = c("Healthy", "Overweight or Obese"))
  ) |>
  filter(height_in >= 55, height_in <= 78, weight_lbs <= 350)

# Scatter plot of height vs. weight colored by BMI status
ggplot(nhanes_bmi, aes(x = height_in, y = weight_lbs, color = status)) +
  geom_point(alpha = 0.45, size = 1.8) +
  scale_color_manual(
    values = c(
      "Healthy" = "#2e7d32",
      "Overweight or Obese" = "#d9534f"
    )
  ) +
  labs(
    title = "Adult Body Mass Index (BMI) by Height and Weight",
    subtitle = "NHANES Survey Participants (Green = Healthy, Red = Overweight or Obese)",
    x = "Height (inches)",
    y = "Weight (lbs)",
    color = "BMI Status"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.5),
    panel.grid.minor = element_blank()
  )
```


## US Adult Obesity Prevalence (1999–2023)

This visualization shows the trend in age-adjusted obesity prevalence ($\text{BMI} \ge 30$) among adults aged 20 and older in the United States from 1999 through 2023, based on data collected by the National Health and Nutrition Examination Survey (NHANES).

```{r}
# CDC NHANES age-adjusted adult obesity prevalence across survey cycles
obesity_trends <- tibble(
  cycle = c(
    "1999–2000", "2001–2002", "2003–2004", "2005–2006",
    "2007–2008", "2009–2010", "2011–2012", "2013–2014",
    "2015–2016", "2017–2018", "2017–2020", "2021–2023"
  ),
  year = c(1999.5, 2001.5, 2003.5, 2005.5, 2007.5, 2009.5, 2011.5, 2013.5, 2015.5, 2017.5, 2018.5, 2022.5),
  prevalence = c(30.5, 30.5, 32.2, 34.3, 33.7, 35.7, 34.9, 37.7, 39.6, 42.4, 41.9, 40.3)
)

# Build plot with ggplot2
p <- ggplot(obesity_trends, aes(x = year, y = prevalence)) +
  geom_line(aes(group = 1), color = "#2b6cb0", linewidth = 1) +
  geom_point(
    aes(text = paste0("Cycle: ", cycle, "<br>Prevalence: ", prevalence, "%")),
    color = "#2b6cb0",
    size = 2.5
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, "%"),
    limits = c(25, 45)
  ) +
  scale_x_continuous(
    breaks = seq(2000, 2022, by = 4),
    limits = c(1998, 2024)
  ) +
  labs(
    title = "Adult Obesity Rate Over Time (NHANES 1999–2023)",
    x = "Survey Period",
    y = "Obesity Prevalence (%)"
  ) +
  theme_minimal()

# Convert to interactive plotly widget
ggplotly(p, tooltip = "text")
```

Between 1999–2000 and 2017–2018, adult obesity prevalence rose steadily from 30.5% to over 42%, before leveling off near 40% in the most recent 2021–2023 cycle.

## Diet-Related Chronic Diseases (1999–2023)

This visualization tracks the age-adjusted prevalence of major chronic health conditions in the U.S. that are strongly influenced by diet, nutrition, and food quality, based on examination data from CDC NHANES.

```{r}
# CDC NHANES age-adjusted chronic disease prevalence across survey cycles
cycles <- c(
  "1999–2000", "2001–2002", "2003–2004", "2005–2006", "2007–2008",
  "2009–2010", "2011–2012", "2013–2014", "2015–2016", "2017–2018",
  "2017–2020", "2021–2023"
)
years <- c(1999.5, 2001.5, 2003.5, 2005.5, 2007.5, 2009.5, 2011.5, 2013.5, 2015.5, 2017.5, 2018.5, 2022.5)

chronic_trends <- tibble(
  cycle = rep(cycles, 4),
  year = rep(years, 4),
  disease = c(
    rep("Hypertension", 12),
    rep("Metabolic Syndrome", 12),
    rep("High Cholesterol", 12),
    rep("Diabetes", 12)
  ),
  rate = c(
    # Hypertension (BP >= 130/80 or medication)
    47.0, 46.0, 44.9, 43.8, 43.2, 42.1, 43.9, 41.7, 43.6, 45.4, 45.1, 45.5,
    # Metabolic Syndrome (cluster of cardiometabolic risk factors)
    28.2, 29.0, 29.8, 30.7, 31.5, 31.9, 33.2, 34.6, 35.8, 37.1, 37.5, 36.8,
    # High Cholesterol (total cholesterol >= 240 mg/dL)
    18.3, 17.2, 16.5, 15.7, 14.6, 13.4, 12.9, 12.1, 12.4, 11.5, 10.8, 10.5,
    # Diabetes (total diagnosed and undiagnosed)
    9.7, 10.4, 10.9, 11.5, 11.9, 12.2, 12.5, 13.0, 14.9, 14.8, 14.8, 14.3
  )
)

# Build plot with ggplot2
p_chronic <- ggplot(chronic_trends, aes(x = year, y = rate, color = disease)) +
  geom_line(aes(group = disease), linewidth = 1) +
  geom_point(
    aes(text = paste0("Condition: ", disease, "<br>Cycle: ", cycle, "<br>Prevalence: ", rate, "%")),
    size = 2.5
  ) +
  scale_color_manual(
    values = c(
      "Hypertension" = "#e65100",        # deep orange
      "Metabolic Syndrome" = "#0288d1",  # blue
      "High Cholesterol" = "#c2185b",    # berry/magenta
      "Diabetes" = "#7b1fa2"             # purple
    )
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, "%"),
    limits = c(5, 55)
  ) +
  scale_x_continuous(
    breaks = seq(2000, 2022, by = 4),
    limits = c(1998, 2024)
  ) +
  labs(
    title = "Prevalence of Diet-Related Chronic Diseases (NHANES 1999–2023)",
    x = "Survey Period",
    y = "Age-Adjusted Prevalence (%)",
    color = "Condition"
  ) +
  theme_minimal()

# Convert to interactive plotly widget
ggplotly(p_chronic, tooltip = "text")
```

### About These Diseases and Their Links to Food

::: {.callout-note collapse="true" icon=false}
## Hypertension (High Blood Pressure)
Defined as blood pressure $\ge 130/80\text{ mmHg}$ or taking blood pressure medication. Excessive sodium (salt) from processed and restaurant foods, combined with diets low in potassium and magnesium from fresh fruits and vegetables, is a key driver.
:::

::: {.callout-note collapse="true" icon=false}
## Metabolic Syndrome
A cluster of at least three concurrent cardiometabolic risk factors: abdominal obesity (large waist circumference), high triglycerides, low HDL ("good") cholesterol, high blood pressure, and elevated fasting blood glucose. Heavily promoted by diets high in ultra-processed foods and refined carbohydrates.
:::

::: {.callout-note collapse="true" icon=false}
## High Total Cholesterol (Hypercholesterolemia)
Defined as blood total cholesterol $\ge 240\text{ mg/dL}$. Elevated by diets high in saturated fats and industrial trans fats. Its long-term decline over this period is largely credited to widespread clinical screening and statin medication use.
:::

::: {.callout-note collapse="true" icon=false}
## Diabetes (Total Type 2 & Type 1)
Defined by elevated blood glucose ($\text{HbA1c} \ge 6.5\%$ or fasting glucose $\ge 126\text{ mg/dL}$), with Type 2 accounting for over 90% of cases. Strongly linked to sugar-sweetened beverages, ultra-processed food intake, and excess caloric consumption that lead to insulin resistance.
:::

:::

## Food Consumption Trends (Ultra-Processed vs. Unprocessed)

This visualization tracks the proportion of daily caloric intake derived from **ultra-processed foods** (red) compared to **unprocessed or minimally processed foods** (green) across all survey cycles from 1999 through 2023 based on NHANES dietary recall data.

```{r}
# NHANES dietary energy intake (% of total kcal) categorized by NOVA food processing level
food_trends <- tibble(
  cycle = rep(
    c(
      "1999–2000", "2001–2002", "2003–2004", "2005–2006", "2007–2008",
      "2009–2010", "2011–2012", "2013–2014", "2015–2016", "2017–2018",
      "2017–2020", "2021–2023"
    ),
    2
  ),
  year = rep(
    c(1999.5, 2001.5, 2003.5, 2005.5, 2007.5, 2009.5, 2011.5, 2013.5, 2015.5, 2017.5, 2018.5, 2022.5),
    2
  ),
  food_type = c(
    rep("Ultra-Processed Food", 12),
    rep("Unprocessed Food", 12)
  ),
  rate = c(
    52.8, 53.5, 54.0, 54.8, 55.2, 55.6, 56.1, 56.4, 56.7, 57.0, 56.5, 53.0,
    33.5, 32.7, 32.1, 31.2, 30.5, 29.8, 29.1, 28.5, 27.9, 27.4, 28.0, 31.2
  )
)

# Build plot with red for processed and green for unprocessed
p_food <- ggplot(food_trends, aes(x = year, y = rate, color = food_type)) +
  geom_line(aes(group = food_type), linewidth = 1) +
  geom_point(
    aes(text = paste0("Type: ", food_type, "<br>Cycle: ", cycle, "<br>Caloric Share: ", rate, "%")),
    size = 2.5
  ) +
  scale_color_manual(
    values = c(
      "Ultra-Processed Food" = "#d9534f", # red
      "Unprocessed Food" = "#2e7d32"       # green
    )
  ) +
  scale_y_continuous(
    labels = \(x) paste0(x, "%"),
    limits = c(20, 65)
  ) +
  scale_x_continuous(
    breaks = seq(2000, 2022, by = 4),
    limits = c(1998, 2024)
  ) +
  labs(
    title = "Ultra-Processed vs. Unprocessed Food Consumption (NHANES 1999–2023)",
    x = "Survey Period",
    y = "Share of Daily Energy Intake (%)",
    color = "Food Category"
  ) +
  theme_minimal()

# Convert to interactive plotly widget
ggplotly(p_food, tooltip = "text")
```

From 1999 to 2018, daily caloric intake from ultra-processed food (red) rose from 52.8% to 57.0%, while unprocessed food (green) dropped from 33.5% to 27.4%, followed by a post-pandemic shift toward whole foods in 2021–2023.
